---
title: "Week 1, Class 1: Practice Exercises"
subtitle: "Introduction to Quantitative Political Analysis"
date: "2025"
format:
  pdf:
    toc: false
    number-sections: true
    colorlinks: true
---

```{r setup, echo=F, include=F}
library(knitr)
library(formatR)
library(tidyverse)
args <- commandArgs(trailingOnly = FALSE)
script <- sub("^--file=", "", args[grep("^--file=", args)])
setwd(dirname(normalizePath(script)))
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

# Non-AI Exercises

## 1. Vocabulary & Concepts

### 1.1 Code Detective

Explain what each line of this code does:

``` r
electoral_votes <- 270
winner <- "Biden"
margin <- 81283501 - 74223975
percentage <- margin / 155507476 * 100
```

Line 1:

Line 2:

Line 3:

Line 4:

## 2. Historical Example: John Snow

### 2.1 The conventional wisdom in 1854 was that cholera spread through:

a)  Contaminated water
b)  Bad air (miasma)
c)  Person-to-person contact
d)  Poor nutrition

Answer:

### 2.2 Data-Driven Decision Making

John Snow challenged the conventional wisdom about cholera transmission. What made his approach "quantitative" rather than just observational? Why was mapping the data crucial to his discovery?

Answer:

## 3. Critical Thinking: AI and Analysis

### 3.1 Critical Thinking with AI

Why is it important to verify AI-generated analysis rather than accepting it automatically? Give an example of how an AI might produce technically correct code that leads to a misleading conclusion.

Answer:

## 4. Applications of Quantitative Political Analysis

# AI Exercises

**Tips for Working with Claude:**

- Ask for **R code using only tidyverse** (no other packages)
- Request **simple, focused answers** to your specific question—not complex analyses
- Ask Claude to **explain what the code is doing** since you're learning
- Avoid asking for visualizations or plots in these exercises
- Include the output of `glimpse()` in your prompt so Claude knows your variable names

**Example prompt:** "Using tidyverse in R, calculate the mean age by party_id. Keep the code simple and explain what each line does. Here is what my data looks like: [paste glimpse output]"

## 5. Introduction to Political Data

**Dataset: nat_pol_attitudes.csv**

**Description**: Simulates a nationally representative survey measuring political attitudes, ideology, and demographics.

**Variables**:

- `respondent_id`: Unique respondent ID (int)
- `age`: Age in years, 18-90 (int)
- `gender`: male, female, nonbinary (factor)
- `race_ethnicity`: White, Black, Latino, Asian, Other (factor)
- `education`: Less than HS, HS, Some College, BA, Postgrad (ordered)
- `income_bracket`: Ten brackets from <$10k to >$200k (ordered)
- `ideology`: 1 (very liberal) to 7 (very conservative) (int)
- `party_id`: Democrat, Republican, Independent, Other (factor)
- `trust_gov`: 0-10 political trust scale (int)
- `policy_support_env`: Support for environmental regulation, 0/1 (binary)
- `policy_support_guns`: Support for stricter gun laws, 0/1 (binary)

### 5.1 Data Exploration

```{r}
# Load the dataset
nat_pol_attitudes <- read_csv("nat_pol_attitudes.csv")

# Explore the structure of the data
glimpse(nat_pol_attitudes)
```

```{r}

```

### 5.2 Basic Summary Statistics

Work with Claude to calculate average age by party affiliation. (Bonus: Try to do this without using Claude.)

```{r}

```

## 6. Understanding Election Data

**Dataset: precinct_elections.csv**

**Description**: Precinct-level election returns with demographics.

**Variables**:

- `state`: Two-letter state abbreviation (factor)
- `county`: County name (string)
- `precinct_id`: Unique precinct identifier (int)
- `year`: Election year (int)
- `reg_voters`: Number of registered voters (int)
- `turnout`: Voter turnout percentage (num)
- `dem_votes`: Democratic candidate votes (int)
- `rep_votes`: Republican candidate votes (int)
- `median_income`: Precinct median household income (num)
- `pct_bachelor`: Percentage with bachelor's degree (num)
- `race_black`: Percentage Black population (num)
- `race_hispanic`: Percentage Hispanic population (num)

### 6.1 Loading and Initial Analysis

```{r}
# Load the dataset
precinct_elections <- read_csv("precinct_elections.csv")

# Examine the data
glimpse(precinct_elections)
```

### 6.2 Calculating Turnout

Ask Claude to help you calculate voter turnout (total votes / registered voters).

```{r}

```

## 7. Congressional Data Analysis

**Dataset: congress_press.csv**

**Description**: Corpus of press releases issued by U.S. legislators.

**Variables**:

- `release_id`: Unique press-release ID (int)
- `member_id`: Legislator ID (int)
- `chamber`: House, Senate (factor)
- `party`: Democrat, Republican, Independent (factor)
- `ideology_score`: DW-NOMINATE first dimension (num)
- `state`: Two-letter abbreviation (factor)
- `date`: Release date (date)
- `topic`: Ten topics like Health, Economy, Foreign Policy, etc. (factor)
- `sentiment_score`: -1 to 1 sentiment scale (num)
- `contains_attack`: Indicator of partisan attack language (binary)

### 7.1 Understanding Press Release Patterns

```{r}
# Load the dataset
congress_press <- read_csv("congress_press.csv")

# Look at the data structure
glimpse(congress_press)
```

### 7.2 Party Differences

Work with Claude to compare how Democrats and Republicans differ in press release topics.

```{r}

```

